Annual Analysis of Delays in TTC Streetcars for the year 2023

### 2024-09-27 paper/paper.qmd
Added citation for professor's starter code and updated sketches
Update data cleaning to convert numbers to numeric type, add details about measurement and cleaned_data update

library(tidyverse)
library(ggplot2)
library(gt)
library(GGally)

@@

These delays make it especially difficult for individuals that commute on a daily basis. Furthermore, one can image the inconveniences caused due to extended delays by streetcars during weather conditions like the winter months. In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced. In this paper, we make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr). For our dataset, we use the provided data for the year 2023 by OpenData Toronto (@opendatatoronto). The reported graphs in our study is generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) library. Code cleanup is performed using the Janitor (@janitor) package.

@@

The following table describes the data types for the selected features in our study from the cleaned dataset.
```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false

@@

# Filter the data for times between 08:00 and 01:00 i.e day shift
avg_delay_day_shift <- data |> 
  group_by(hour) |>
  summarise(average_delay = mean(min_delay)) |>
  filter(hour >= "08" | hour <= "01")

ggplot(avg_delay_day_shift, aes(x = as.numeric(hour), y = average_delay)) +
  geom_line(color = "dodgerblue1", size = 1, alpha = 0.5) +
  geom_point(color = "dodgerblue4", size = 2) +
  geom_text(aes(label = round(average_delay, 1)), 
            nudge_y = 0.25, 
            size = 1.75) + 
  ggtitle("Average delay in minutes for morning shift") +
  xlab("Hour of Day") +
  ylab("Average Delay (Minutes)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```
@fig-delay-dayshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the day. We observe that there is high variability in the delays throughout the day with the minimum delay being about 13.78 minutes and the maximum being roughly 20 minutes.

@@

# Filter the data for times between 02:00 and 07:00 i.e night shift
avg_delay_night_shift <- data |> 
  group_by(hour) |>
  filter(hour >= "02" & hour <= "07") |>
  summarise(average_delay = mean(min_delay))

# Plot the graph with the reordered x-axis
ggplot(avg_delay_night_shift, aes(x = as.numeric(hour), y = average_delay)) +
  geom_line(color = "red1", size = 1, alpha = 0.5) +
  geom_point(color = "red3", size = 2) +
  geom_text(aes(label = round(average_delay, 1)), 
            vjust = 2.5, 
            size = 1.75) + 
  ggtitle("Average delay in minutes for night shift") +
  xlab("Hour of Day") +
  ylab("Average Delay (Minutes)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```
@fig-delay-nightshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the night. We observe that there is decreasing trend in the delays throughout the night after 3 AM with the minimum delay being about 14 minutes and the maximum roughly 27 minutes.

>>>>

library(tidyverse)
library(ggplot2)
library(gt)

@@

These delays make it especially difficult for individuals that commute on a daily basis. Furthermore, one can image the inconveniences caused due to extended delays by streetcars during weather conditions like the winter months. In this paper, we aim to dive deeper into the cause of delays in TTC, how these vary by the times of day and the various seasons and what incidents are the most occurring for the delays experienced. The starter code for this paper was provided by Professor Rohan (@citeRohan). We make use R (@citeR) and its supported packages like tidyverse  (@tidyverse) and dplyr (@dplyr). For our dataset, we use the provided data for the year 2023 by OpenData Toronto (@opendatatoronto). The reported graphs in our study is generated using ggplot (@ggplot2) and the tables are generated using the gt (@gt) library. Code cleanup is performed using the Janitor (@janitor) package.

@@

The data is collected and maintained by TTC and is updated monthly as reported by the website. Since the data is stored as a CSV, several numerical fields like the streetcar line or delay in minutes are saved as characters instead of numbers. Fields like date are saved as appropriate format of datetime type.

The dataset was last updated September 24, 2024 ensuring consistent updates to the collected data. OpenData Toronto also provides functionality and sufficient instructions for developers to easily query their data using scripts in Python, R and Node.js instead of having to manually download them.

The website doesn't provided information on how the data was sourced but we can assume that the data is recorded by taking the difference in the reported arrival time via the TTC app and the actual time the streetcar arrives at various stops. The collected data represents common types like characters, numeric and datetime. Therefore, for the purposes of this study, the procedure and type of the data reported is appropriate for our use and we do not require any extra information to make any assumptions about the data types.

The following table describes the data types for the selected features in our study from the cleaned dataset after the necessary transformations.
```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false

@@

# Filter the data for times between 08:00 and 00:00 i.e day shift
avg_delay_day_shift <- data |> 
  group_by(hour) |>
  summarise(average_delay = mean(min_delay)) |>
  filter(hour >= 8 & hour <= 23)

# Plot the graph
ggplot(avg_delay_day_shift, aes(x = as.numeric(hour), y = average_delay)) +
  geom_line(color = "dodgerblue1", size = 1, alpha = 0.5) +
  geom_point(color = "dodgerblue4", size = 2) +
  geom_text(aes(label = round(average_delay, 1)), 
            nudge_y = 0.25, 
            size = 1.75) + 
  ggtitle("Average delay in minutes for morning shift") +
  xlab("Hour of Day") +
  ylab("Average Delay (Minutes)") +
  scale_x_continuous(breaks = seq(8, 23)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```
@fig-delay-dayshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the day. We observe that there is high variability in the delays throughout the day with the minimum delay being about 12.2 minutes at 8 AM and the maximum being roughly 17 minutes at 11 PM.

@@

# Filter the data for times between 00:00 and 07:00 i.e night shift
avg_delay_night_shift <- data |> 
  group_by(hour) |>
  filter(hour >= 0 & hour <= 7) |>
  summarise(average_delay = mean(min_delay))

# Plot the graph
ggplot(avg_delay_night_shift, aes(x = hour, y = average_delay)) +
  geom_line(color = "red1", size = 1, alpha = 0.5) +
  geom_point(color = "red3", size = 2) +
  geom_text(aes(label = round(average_delay, 1)), 
            vjust = 2.5, 
            size = 1.75) + 
  ggtitle("Average delay in minutes for night shift") +
  xlab("Hour of Day") +
  ylab("Average Delay (Minutes)") +
  scale_x_continuous(breaks = seq(0, 7)) + 
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for clarity
  )
```
@fig-delay-nightshift highlights the annual average delay in minutes experienced by the hour of day for streetcars that run during the night. We observe that there is maximum delay at 3 AM for roughly 25 minutes and the minimum delay being about 11.4 minutes at 7 AM.
